---
title: "Latent time joint mixed effect model simulation & fit"
author: "Dan Li and  Michael C. Donohue"
output: 
  rmarkdown::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Latent time joint mixed effect model simulation and fit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr_settings, results = 'hide', echo = FALSE, message = FALSE, cache = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, tidy = FALSE, 
  fig_width = 6.5, fig_height = 4, comment = NA)
```

```{r libraries}
library(dplyr)
library(tidyr)
library(ltjmm)
library(rstan)
```

```{r settings, echo=FALSE}
theme_set(theme_bw(base_size = 12))
cbbPalette <- c("#0072B2", "#D55E00", "#E69F00", 
                "#F0E442", "#56B4E9", "#CC79A7", "#009E73", "#999999", "#000000")
scale_colour_discrete <- function(...) scale_colour_manual(..., values=cbbPalette)
scale_fill_discrete <- function(...) scale_fill_manual(..., values=cbbPalette)
```

This vignettes demonstrates simulating and fitting latent time joint mixed effect models as described in Li, et al. 2017\footnote{Li, D., Iddi, S., Thompson, W. K., Donohue, M. C., for the Alzheimer's Disease Neuroimaging Initiative. (2017). Bayesian latent time joint mixed effect models for multicohort longitudinal data. \emph{Statistical methods in medical research}, \url{https://doi.org/10.1177/0962280217737566}.}.

The simulated data includes latent times. Random intercepts and slopes are simulated 

# Simulate one example dataset
```{r simulationParameters}
rng_seed <- 20161001
set.seed(rng_seed)
n <- 400   # subjects
p <- 4     # outcomes
t <- 4     # time points

subjects <- data.frame(id = 1:n,
  age.0 = rnorm(n, 75, 5),
  apoe = sample(c(0, 1), size = n, replace = TRUE))

subject_time_outcome <- expand.grid(id = 1:n, visit = c(1, 2, 3, 4), outcome = 1:p)
subject_time_outcome <- subject_time_outcome[order(subject_time_outcome$id, 
  subject_time_outcome$outcome),]

for(i in 1:n){
  time.year <- sort(runif(t, 0, 10))  # t number of time points
  for(j in 1:p){
    subject_time_outcome$year[(subject_time_outcome$id == i) & 
      (subject_time_outcome$outcome == j)] <- time.year
  }
}

dd <- right_join(subjects, subject_time_outcome, by = 'id')
dd$Y <- NA
dd0 <- dd
setup <- ltjmm(Y ~ year | 1 | id | outcome, data = dd)

## variance parameters 
sigma_y <- c(0.1, 0.2, 0.3, 0.25)
# one less degree of freedom for intercepts due to identifiability constraint:
sigma_alpha0 <- c(0.5, 1, 0.8) 
sigma_alpha1 <- c(1, 2, 1.5, 1)
sigma_alpha <- c(sigma_alpha0, sigma_alpha1)
sigma_delta <- 4
N_X <- 1
beta <- matrix(c(1, 0.5, 2, 0.8), p, N_X)
gamma <- c(0.2, 0.1, 0.25, 0.5)

dd$id <- as.factor(dd$id)
dd$Outcome <- factor(dd$outcome, levels = 1:p, labels = paste0('Y', 1:p))
dd$APOE <- factor(dd$apoe, levels = 0:1, labels = c('-', '+'))

dd2 <- simulate(setup,
  beta = beta,
  gamma = gamma,
  sigma_diag = sigma_alpha,
  sigma_delta = sigma_delta,
  sigma_y = sigma_y,
  seed = 201610014)

dd$Y <- dd2$y
dd$Q <- dd$Z <- dd$Z2 <- NA
for(oc in unique(dd$outcome)){
  subs <- dd$outcome == oc
  ecdf.fun <- ecdf(dd$Y[subs])
  dd$Q[subs] <- ecdf.fun(dd$Y[subs])
  dd$Z2[subs] <- qnorm(dd$Q[subs])
  dd$Z[subs] <- scale(dd$Y[subs])
}
```

# Fit LTJMM with independent random effects with Stan
```{r stanFit, eval=FALSE}
fit <- ltjmm_stan(Y ~ year |
    1  | # fixed effects direct on outcome
    id | outcome,
  data = dd,
  pars = c('beta', 'delta', 'alpha0', 'alpha1', 'gamma', 
    'sigma_alpha0', 'sigma_alpha1', "sigma_delta", 'sigma_y', 'log_lik'),
  open_progress = FALSE, chains = 2, iter = 1000, thin = 2, cores = 2,
  seed = rng_seed)
save(fit, file = 'sim_results.rdata')
```

```{r loadResults}
load(file = 'sim_results.rdata')
fit.sum <- summary(fit)$summary
```

## Diagnostic plots
```{r traceplotBeta, fig_width = 6.5, fig_height = 4}
plot(fit, pars='beta', inc_warmup=TRUE, plotfun='trace')
```

```{r traceplotGamma}
plot(fit, pars='gamma', inc_warmup=TRUE, plotfun='trace')
```

```{r traceplotSigmaDelta}
plot(fit, pars='sigma_delta', inc_warmup=TRUE, plotfun='trace')
```

## True versus posterior mean of latent time shifts
```{r biasDeltaPlot}
delta <- dd2$delta
delta.posteriormean <- fit.sum[(grepl('delta', rownames(fit.sum))) & 
  (!(grepl('sigma_delta', rownames(fit.sum)))), 1]
par(mgp = c(2.2, 0.45, 0), tcl = -0.4, mar = c(3.3, 3.6, 1.1, 1.1))
plot(delta, delta.posteriormean,
  xlim = range(delta), ylim = range(delta), 
  xlab = expression(paste("True value of time shift ", delta[i])),
  ylab = expression(paste("Posterior mean of time shift ", delta[i])))
abline(0, 1, lwd=2, col='red', lty = 2)
```

## True versus posterior mean of random intercepts and slopes for each outcome
```{r biasAlphaPlot}
alpha0true <- as.data.frame(dd2$alpha0) %>% mutate(id = 1:n, parameter='alpha0') %>% 
  gather(outcome, truth, V1:V4)
alpha1true <- as.data.frame(dd2$alpha1) %>% mutate(id = 1:n, parameter='alpha1') %>% 
  gather(outcome, truth, V1:V4)
alphapm <- data.frame(
    parameter.id.outcome = grep('alpha', rownames(fit.sum), value=TRUE), 
    posterior.mean = fit.sum[grepl('alpha', rownames(fit.sum)), 1]) %>% 
  separate(parameter.id.outcome, c('parameter', 'id', 'outcome', 'other')) %>% 
  mutate(id = as.numeric(id))

pd <- full_join(alpha0true, alpha1true, by=c('id', 'outcome', 'parameter', 'truth')) %>%
  mutate(outcome = gsub('V', '', outcome)) %>% 
  full_join(alphapm) 
ggplot(pd, aes(x=truth, y=posterior.mean)) + 
  geom_point(aes(shape=parameter, color=parameter), alpha=0.25) +
  geom_abline(intercept=0, slope=1) + facet_wrap(~outcome, scales='free')
```

# Fit LTJMM with multivariate random effects with Stan
```{r stanFitMulti, eval=FALSE}
fit <- ltjmm_stan(Y ~ year |
    1  | # fixed effects direct on outcome
    id | outcome,
  random_effects = "multivariate",
  data = dd,
  pars = c('beta', 'alpha0', 'alpha1', 'gamma', 
    'delta', 'sigma_y', 'log_lik'),
  open_progress = FALSE, chains = 2, iter = 1000, thin = 2, cores = 2)
save(fit, file = 'sim_lt_multi_results.rdata')
```

```{r loadResultsMulti}
load(file = 'sim_lt_multi_results.rdata')
fit.sum <- summary(fit)$summary
```

## Diagnostic plots
```{r traceplotBetaMulti}
plot(fit, pars='beta', inc_warmup=TRUE, plotfun='trace')
```

```{r traceplotGammaMulti}
plot(fit, pars='gamma', inc_warmup=TRUE, plotfun='trace')
```

## True versus posterior mean of latent time shifts
```{r biasDeltaPlotMulti}
delta <- dd2$delta
delta.posteriormean <- fit.sum[(grepl('delta', rownames(fit.sum))) & 
  (!(grepl('sigma_delta', rownames(fit.sum)))), 1]
par(mgp = c(2.2, 0.45, 0), tcl = -0.4, mar = c(3.3, 3.6, 1.1, 1.1))
plot(delta, delta.posteriormean,
  xlim = range(delta), ylim = range(delta), 
  xlab = expression(paste("True value of time shift ", delta[i])),
  ylab = expression(paste("Posterior mean of time shift ", delta[i])))
abline(0, 1, lwd=2, col='red', lty = 2)
```

## True versus posterior mean of random intercepts and slopes for each outcome
```{r biasAlphaPlotMulti}
alpha0true <- as.data.frame(dd2$alpha0) %>% mutate(id = 1:n, parameter='alpha0') %>% 
  gather(outcome, truth, V1:V4)
alpha1true <- as.data.frame(dd2$alpha1) %>% mutate(id = 1:n, parameter='alpha1') %>% 
  gather(outcome, truth, V1:V4)
alphapm <- data.frame(
    parameter.id.outcome = grep('alpha', rownames(fit.sum), value=TRUE), 
    posterior.mean = fit.sum[grepl('alpha', rownames(fit.sum)), 1]) %>% 
  separate(parameter.id.outcome, c('parameter', 'id', 'outcome', 'other')) %>% 
  mutate(id = as.numeric(id)) %>% 
  filter(parameter != 'sigma')

pd <- full_join(alpha0true, alpha1true, by=c('id', 'outcome', 'parameter', 'truth')) %>%
  mutate(outcome = gsub('V', '', outcome)) %>% 
  full_join(alphapm) 
ggplot(pd, aes(x=truth, y=posterior.mean)) + 
  geom_point(aes(shape=parameter, color=parameter), alpha=0.25) +
  geom_abline(intercept=0, slope=1) + facet_wrap(~outcome, scales='free')
```

# Fit joint mixed effect model (JMM) with multivariate random effects with Stan
```{r stanFitJMM, eval=FALSE}
fit <- ltjmm_stan(Y ~ year |
    1  | # fixed effects direct on outcome
    id | outcome,
  lt = FALSE,
  random_effects = "multivariate",
  data = dd,
  pars = c('beta', 'alpha0', 'alpha1', 'gamma', 
    'sigma_y', 'log_lik'),
  open_progress = FALSE, chains = 2, iter = 500, 
    warmup = 250, thin = 1, cores = 2)
save(fit, file = 'sim_jmm_results.rdata')
```

```{r loadResultsJMM}
load(file = 'sim_jmm_results.rdata')
fit.sum <- summary(fit)$summary
```

## Diagnostic plots
```{r traceplotBetaJMM}
plot(fit, pars='beta', inc_warmup=TRUE, plotfun='trace')
```

## True versus posterior mean of random intercepts and slopes for each outcome
```{r biasAlphaPlotJMM}
alpha0true <- as.data.frame(dd2$alpha0) %>% mutate(id = 1:n, parameter='alpha0') %>% 
  gather(outcome, truth, V1:V4)
alpha1true <- as.data.frame(dd2$alpha1) %>% mutate(id = 1:n, parameter='alpha1') %>% 
  gather(outcome, truth, V1:V4)
alphapm <- data.frame(
    parameter.id.outcome = grep('alpha', rownames(fit.sum), value=TRUE), 
    posterior.mean = fit.sum[grepl('alpha', rownames(fit.sum)), 1]) %>% 
  separate(parameter.id.outcome, c('parameter', 'id', 'outcome', 'other')) %>% 
  mutate(id = as.numeric(id))

pd <- full_join(alpha0true, alpha1true, by=c('id', 'outcome', 'parameter', 'truth')) %>%
  mutate(outcome = gsub('V', '', outcome)) %>% 
  full_join(alphapm) 
ggplot(pd, aes(x=truth, y=posterior.mean)) + 
  geom_point(aes(shape=parameter, color=parameter), alpha=0.25) +
  geom_abline(intercept=0, slope=1) + facet_wrap(~outcome, scales='free')
```

# Fit mixed effect model (MM) with independent univariate random effects with Stan
```{r stanFitMM, eval=FALSE}
fit <- ltjmm_stan(Y ~ year |
    1  | # fixed effects direct on outcome
    id | outcome,
  lt = FALSE,
  random_effects = "univariate",
  data = dd,
  pars = c('beta', 'alpha0', 'alpha1', 'gamma', 
    'sigma_y', 'log_lik'),
  open_progress = FALSE, chains = 2, iter = 500, 
    warmup = 250, thin = 1, cores = 2)
save(fit, file = 'sim_mm_results.rdata')
```

```{r loadResultsMM}
load(file = 'sim_mm_results.rdata')
fit.sum <- summary(fit)$summary
```

## Diagnostic plots
```{r traceplotBetaMM}
plot(fit, pars='beta', inc_warmup=TRUE, plotfun='trace')
```

```{r traceplotGammaMM}
plot(fit, pars='gamma', inc_warmup=TRUE, plotfun='trace')
```

## True versus posterior mean of random intercepts and slopes for each outcome
```{r biasAlphaPlotMM}
alpha0true <- as.data.frame(dd2$alpha0) %>% mutate(id = 1:n, parameter='alpha0') %>% 
  gather(outcome, truth, V1:V4)
alpha1true <- as.data.frame(dd2$alpha1) %>% mutate(id = 1:n, parameter='alpha1') %>% 
  gather(outcome, truth, V1:V4)
alphapm <- data.frame(
    parameter.id.outcome = grep('alpha', rownames(fit.sum), value=TRUE), 
    posterior.mean = fit.sum[grepl('alpha', rownames(fit.sum)), 1]) %>% 
  separate(parameter.id.outcome, c('parameter', 'id', 'outcome', 'other')) %>% 
  mutate(id = as.numeric(id))

pd <- full_join(alpha0true, alpha1true, by=c('id', 'outcome', 'parameter', 'truth')) %>%
  mutate(outcome = gsub('V', '', outcome)) %>% 
  full_join(alphapm) 
ggplot(pd, aes(x=truth, y=posterior.mean)) + 
  geom_point(aes(shape=parameter, color=parameter), alpha=0.25) +
  geom_abline(intercept=0, slope=1) + facet_wrap(~outcome, scales='free')
```
